Approximately Exact Calculations for Linear Mixed Models

### 2015-05-31 ms.tex
Added run times for code

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
\begin{equation*}
	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 15 iterations there is a total of 9490 boxes, which are displayed in Figure~\ref{fig:hmo1}.

@@

Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation in under 10 seconds.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.  Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

@@

For a Bayesian analysis there are 44 lines, as shown in Figure~\ref{fig:hmoBayeslines}.  Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line for $j = s_z+2$ and the position of the vertical line for $j = s_z+1$ is slightly shifted.
\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmolines_HH11_Bayes.pdf}
	\caption{The 44 lines, $j=1, 2, \dots, 44$, for the Bayesian analysis of the HMO data.
	              Lines 43 and 44 are dashed.}
	\label{fig:hmoBayeslines}
\end{figure}
Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with control parameters
\begin{equation*}
	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
\end{equation*}
The algorithm finished in 22 iterations.  Table~\ref{table:hmo_HH11Bayes} shows the output.
\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 16 & 0 & -1313.42\\
3 & 52 & 3 & -1285.05\\
4 & 132 & 22 & -1270.76\\
5 & 264 & 88 & -1263.98\\
6 & 628 & 195 & -1260.47\\
7 & 1780 & 378 & -1258.71\\
8 & 4188 & 1111 & -1257.70\\
9 & 5676 & 3880 & -1256.88\\
10 & 5136 & 8272 & -1255.56\\
11 & 5528 & 12026 & -1254.20\\
12 & 8236 & 15495 & -1252.84\\
13 & 12488 & 20609 & -1251.50\\
14 & 23852 & 27134 & -1250.26\\
15 & 22740 & 45301 & -1249.22\\
16 & 43228 & 57234 & -1248.77\\
17 & 160208 & 60410 & -1248.77\\
18 & 331900 & 137643 & -1248.77\\
19 & 575196 & 325744 & -1248.77\\
20 & 942860 & 665225 & -1248.77\\
21 & 722412 & 1427482 & -1248.77\\
22 & 0 & 2149894 & -1248.77\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 22 iterations for the Bayesian analysis of the HMO data.}
\label{table:hmo_HH11Bayes}
\end{table}
The Bayesian analysis in Figure~\ref{fig:hmoBayes} can be compared to the $\logRLorig$ analysis in Figure~\ref{fig:hmo1}.  The peak of the log posterior is near $(600, .05)$, very far from the peak of the $\logRLorig$ around $(500,100)$.  The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.048.  The region near $(500,100)$, though having a lower posterior density, is part of a nearly flat plateau covering a large area.  The MCMC draws depicted in \cite{henn&hodges:2014} Figure~2c show that the plateau has significant posterior mass.
%Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.  With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.

>>>>

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
\begin{equation*}
	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$. The run finished in less than 10 seconds.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 15 iterations there is a total of 9490 boxes, which are displayed in Figure~\ref{fig:hmo1}.

@@

Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.  Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

@@

For a Bayesian analysis there are 44 lines, as shown in Figure~\ref{fig:hmoBayeslines}.  Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line for $j = s_z+2$ and the position of the vertical line for $j = s_z+1$ is slightly shifted.
\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmolines_HH11_Bayes.pdf}
	\caption{The 44 lines, $j=1, 2, \dots, 44$, for the Bayesian analysis of the HMO data.
	              Lines 43 and 44 are dashed.}
	\label{fig:hmoBayeslines}
\end{figure}
Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with control parameters
\begin{equation*}
	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
\end{equation*}
The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output.
\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 16 & 0 & -1313.42\\
3 & 52 & 3 & -1285.05\\
4 & 132 & 22 & -1270.76\\
5 & 264 & 88 & -1263.98\\
6 & 628 & 195 & -1260.47\\
7 & 1780 & 378 & -1258.71\\
8 & 4188 & 1111 & -1257.70\\
9 & 5676 & 3880 & -1256.88\\
10 & 5136 & 8272 & -1255.56\\
11 & 5528 & 12026 & -1254.20\\
12 & 8236 & 15495 & -1252.84\\
13 & 12488 & 20609 & -1251.50\\
14 & 23852 & 27134 & -1250.26\\
15 & 22740 & 45301 & -1249.22\\
16 & 43228 & 57234 & -1248.77\\
17 & 160208 & 60410 & -1248.77\\
18 & 331900 & 137643 & -1248.77\\
19 & 575196 & 325744 & -1248.77\\
20 & 942860 & 665225 & -1248.77\\
21 & 722412 & 1427482 & -1248.77\\
22 & 0 & 2149894 & -1248.77\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 22 iterations for the Bayesian analysis of the HMO data.}
\label{table:hmo_HH11Bayes}
\end{table}
The Bayesian analysis in Figure~\ref{fig:hmoBayes} can be compared to the $\logRLorig$ analysis in Figure~\ref{fig:hmo1}.  The peak of the log posterior is near $(600, .05)$, very far from the peak of the $\logRLorig$ around $(500,100)$.  The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.048.  The region near $(500,100)$, though having a lower posterior density, is part of a nearly flat plateau covering a large area.  The MCMC draws depicted in \cite{henn&hodges:2014} Figure~2c show that the plateau has significant posterior mass.
%Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.  With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.
